---
title: "Course Project 2"
author: "Alex Fennell"
output: html_document
---

```{r libraries, message=FALSE,warning=FALSE}
    library(dplyr)
    library(ggplot2)
    library(tidyverse)
    library(stringdist)
## As a note, for tabulizer, the user needs to have Java installed as the same 
#bit format as their R version (e.g. both 64 bit)
    library(tabulizer)
    library(DT)
    library(RColorBrewer)
```

## Data Processing

```{r download-data, cache=TRUE}
fileurl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("CP2data.csv.bz2")){
    download.file(fileurl,"CP2data.csv.bz2",method="curl")
}
storm<-read.csv("CP2data.csv.bz2",stringsAsFactors = FALSE)
```

Since we are interested in the public health and economic consequences of various
event types, the first thing I want to do is select rows that have some amount of
property damage, crop damage, fatalities, or injuries. I also change the date
to a usable format and extract the year information so I can more easily examine
data from particular years. I also remove event types of "?" since that is not
a useful category.
```{r processing the data,cache=TRUE}
stormfilt<-storm%>%
    filter(PROPDMG>0 | CROPDMG>0 | FATALITIES>0 | INJURIES>0)

    stormfilt$BGN_DATE<-strptime(stormfilt$BGN_DATE,format="%m/%d/%Y")
    stormfilt$BGN_DATE<-as.POSIXct(stormfilt$BGN_DATE)
    stormfilt$EVTYPE<-toupper(stormfilt$EVTYPE)
    stormfilt$Year<-format(stormfilt$BGN_DATE,format="%Y")
    stormfilt<-stormfilt[stormfilt$EVTYPE!="?",]
```

Now I am going to download the list of event types (EVTYPES) from the following
link. Using this list I will then match up the official event types with those
in the storm data set. These event types are listed on page 6 of the document
under "2.1.1 Storm Data Event Table". The locate areas function brings up a UI
showing the sixth page of the pdf which the user can use to select the area around
the table of text. The extract tables function will then use these dimensions in
the area argument to pull items from there. I have saved the dimensions in a list
for the sake of reproducibility.
```{r Clean Event Types, cache=TRUE}
fileurl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
#x<-locate_areas(fileurl,pages=6)
x<-list(c(197.6,67.6,564.6,524))
stormeventtypes<-extract_tables(fileurl,output="data.frame",pages=6,
                                area=x)
stormET<-c(stormeventtypes[[1]][,1],stormeventtypes[[1]][,3])
#Remove entries that are not event types
stormET<-stormET[-c(1,2,27,28)]
stormET<-toupper(stormET)
#used jaro-winkler distance (jw) since there are many entries in which there are
#abbreviations and extra words unrelated to categories. This method worked better
#than the default, and resolved NA matches that occurred with the default
stormfilt$EVTYPECOND<-stormET[amatch(stormfilt$EVTYPE,stormET,method="jw",maxDist = 7)]
stormfilt$EVTYPECOND<-as.factor(stormfilt$EVTYPECOND)
```

The next step is to use the CROPDMGEXP and PROPDMGEXP to adjust the actual crop
and property damage totals. There are a few entries in these that only change the 
damage value by a trivial amount (hundreds of dollars) and these will be removed
from the data set, since they will not have a large impact on the analysis as 
damage is typically in the millions or billions. These entries also only account
for a couple hundred observations.
```{r Update damage costs}
stormfilt$PROPDMGEXP <- toupper(stormfilt$PROPDMGEXP) 
stormfilt$CROPDMGEXP <- toupper(stormfilt$CROPDMGEXP)
#Instances of *EXP I am looking to keep
modifier<-c("M","K","B","H","")
stormfilt<-filter(stormfilt,CROPDMGEXP %in% modifier & PROPDMGEXP %in% modifier)
stormfilt<-stormfilt %>%
                mutate(.,PROPDMGFULL= with(.,case_when(
                    (PROPDMGEXP=="H") ~ PROPDMG*100,
                    (PROPDMGEXP=="K") ~ PROPDMG*1000,
                    (PROPDMGEXP=="M") ~ PROPDMG*10^6,
                    (PROPDMGEXP=="B") ~ PROPDMG*10^9,
                    TRUE ~ PROPDMG
                )))
                          
stormfilt<-stormfilt %>%
                mutate(.,CROPDMGFULL= with(.,case_when(
                    (CROPDMGEXP=="H") ~ CROPDMG*100,
                    (CROPDMGEXP=="K") ~ CROPDMG*1000,
                    (CROPDMGEXP=="M") ~ CROPDMG*10^6,
                    (CROPDMGEXP=="B") ~ CROPDMG*10^9,
                    TRUE ~ CROPDMG
                )))
```

# Results

The data has now been cleaned and is ready to be examined. The first question
to be addressed is which types of events are most harmful to population health.
There are two variables in the data set, Fatalities and Injuries which will act
as measures of population health. We will begin our analysis by examining the
total fatalities and injuries caused by various weather event types. I am using
total because this gives the best indication of the severity of the weather event.
This will then help to filter out which event types are of the most interest.
I am also only going to examine data from 1996 and on, since this is when all 
categories of event types were collected. Otherwise the data would be extremely
skewed to reflect data on event types that have been collected since the 
beginning (e.g. Tornadoes)

```{r Population health analysis}
totalstats<-stormfilt %>%
    filter(Year>"1995" & (FATALITIES>0 | INJURIES>0 )) %>%
    group_by(EVTYPECOND) %>%
    summarise(Total_Fatalities=sum(FATALITIES),Total_Injuries=sum(INJURIES))
#datatable(totalstats,class="cell-border stripe")
```
### Data Selection
The results from this analysis show that there is a wide range in total 
fatalities/injuries from single digits to thousands. Based on this, lets use 
event types that are above the 75 percentile for total fatalities and injuries

```{r quantile data selection}
#Establishing quantle cutoff values
quant<-totalstats %>%
        summarise(quantFAT=quantile(Total_Fatalities,probs=.75),
              quantINJ=quantile(Total_Injuries,probs=.75))

Eventnames<-totalstats$EVTYPECOND[totalstats$Total_Fatalities>quant$quantFAT & 
               totalstats$Total_Injuries>quant$quantINJ]

healthdata<-totalstats%>%
    filter(EVTYPECOND %in% Eventnames)%>%
    arrange(desc(Total_Fatalities))
    
healthdata$EVTYPECOND<-factor(healthdata$EVTYPECOND,levels=as.character(healthdata$EVTYPECOND))

```



```{r Tidytext function}
#### This function is part of the TIDYTEXT package authored by Julia Silge
#### and David Robinson. The package was not downloading this function
#### so I used the code they have documented in the function
#### https://rdrr.io/github/juliasilge/tidytext/src/R/reorder_within.R

reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  if (!is.list(within)) {
    within <- list(within)
  }

  new_x <- do.call(paste, c(list(x, sep = sep), within))
  stats::reorder(new_x, by, FUN = fun)
}

```

### Plots Demonstrating harmful effects of Severe Weather events on Population Health

```{r Health plots}
healthdata %>%
     gather(key,value,-EVTYPECOND) %>%
    ggplot(aes(x=reorder_within(EVTYPECOND,-value,key),y=value,
                           fill=EVTYPECOND))+
     geom_bar(stat="identity",position="dodge")+
    facet_wrap(~key,scales="free")+
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.x = element_blank(),
          plot.title = element_text(hjust=.5),
          legend.position = "bottom")+
    scale_fill_brewer(palette = "Set1")+
    labs(y="Total Number of Incidents",
         title = "Total Fatalities and Injuries from Weather from 1996-2011")+
    guides(fill=guide_legend(title="Weather Event"))

```

#### Plot Summary

These plots show the total amount of fatalities and injuries as a result of 8 
weather events from 1996-2011. These weather events were chosen as they were in
the 75th percentile of fatalities and injuries.
Notice the scale between fatalities and injuries differs by about
a factor of 10 (Many more injuries than fatalities). Examining the Fatalities plot
shows that Excessive Heat, and Tornado are the largest contributors to 
fatalities, resulting in more than one thousand fatalities. In the total injuries
plot, tornado, flood, and excessive heat produce that largest number of injuries.
Based on this, Tornadoes seem to be the most harmful to population health as they
result in large amounts of fatalities and massive amounts of injuries. Excessive
heat is the next largest threat to health causing a large amount of fatalities and
injuries. The hierarchy of weather events from here is less clear, but all the
other weather events result in hundreds of fatalities and thousands of injuries.


## Economic Consequences of Weather Events

Now we will do an analysis focusing on the economic impact of weather events. 
This will be very similar in scope to the previous analysis, focusing on events
after 1995 and including events that caused either crop damage, or property damage.
Then we will select events that caused the most property damage as selected by the 
75th percentile.
```{r Population economic analysis}
totalcost<-stormfilt %>%
    filter(Year>"1995" & (PROPDMGFULL>0 | CROPDMGFULL>0 )) %>%
    group_by(EVTYPECOND) %>%
    summarise(Total_Property_Damage=sum(PROPDMGFULL),
              Total_Crop_Damage=sum(CROPDMGFULL))

#Establishing quantle cutoff values
quantcost<-totalcost %>%
        summarise(quantPROP=quantile(Total_Property_Damage,probs=.75),
              quantCROP=quantile(Total_Crop_Damage,probs=.75))

Eventnames<-totalcost$EVTYPECOND[totalcost$Total_Property_Damage>quantcost$quantPROP & 
               totalcost$Total_Crop_Damage>quantcost$quantCROP]

costdata<-totalcost%>%
    filter(EVTYPECOND %in% Eventnames)%>%
    arrange(desc(Total_Property_Damage))
    
costdata$EVTYPECOND<-factor(costdata$EVTYPECOND,levels=as.character(costdata$EVTYPECOND))

```


### Plots demonstrating harmful economic effects of Severe Weather events 

```{r Cost plots}
costdata %>%
     gather(key,value,-EVTYPECOND) %>%
    ggplot(aes(x=reorder_within(EVTYPECOND,-value,key),y=value,
                           fill=EVTYPECOND))+
     geom_bar(stat="identity",position="dodge")+
    facet_wrap(~key,scales="free")+
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.x = element_blank(),
          plot.title = element_text(hjust=.5),
          legend.position = "bottom")+
    scale_fill_brewer(palette = "Set2")+
    labs(y="Total Cost (in Dollars)",
         title = "Total Crop and property damage from Weather from 1996-2011")+
    guides(fill=guide_legend(title="Weather Event"))

```
