---
title: "The Effect of Weather Events on Population Health and Economy"
author: "Alex Fennell"
output: html_document
---
# Synopsis
In this report I examine the economic and public health consequences of severe 
weather events. The data is obtained from the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. The data includes records from 1950-2011 which
contain information such as where and when the weather events occur as well as
information on fatalities, injuries, and property damage. I narrowed the search
to include the years between 1996-2011 as this was when all the currently 
available weather events were recorded. Based on this time window, tornadoes were
the most dangerous weather event in relation to public health followed by excessive
heat and floods. In relation to the economic consequences, hurricanes and flooding
were the most costly weather events, followed by flash floods and hail.

```{r libraries, message=FALSE,warning=FALSE}
    library(tidyverse)
    library(stringdist)
## As a note, for tabulizer, the user needs to have Java installed as the same 
#bit format as their R version (e.g. both 64 bit)
    library(tabulizer)
    library(DT)
    library(RColorBrewer)
```

## Data Processing
The data is obtained from the NOAA storm database. I first download the file and
then use the read csv function as it can handle .bz2 files. I use stringsAsFactors
= FALSE since I need to clean up the data and it will be easier to do that when
the data is formatted as strings.
```{r download-data, cache=TRUE,warning=FALSE}

fileurl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("CP2data.csv.bz2")){
    download.file(fileurl,"CP2data.csv.bz2",method="curl")
}
storm<-read.csv("CP2data.csv.bz2",stringsAsFactors = FALSE)
```

Now I can do a quick check of the data to make sure there are the correct number
of observations and columns (902297 and 37 respectively) and its structure.
```{r}
str(storm)
```

Since I am interested in the public health and economic consequences of various
event types, the first thing I want to do is select rows that have some amount of
property damage, crop damage, fatalities, or injuries. I also change the date
to a usable format and extract the year information so I can more easily examine
data from particular years. I also remove event types of "?" since that is not
a useful category.
```{r processing the data,cache=TRUE}
stormfilt<-storm%>%
    filter(PROPDMG>0 | CROPDMG>0 | FATALITIES>0 | INJURIES>0)

    stormfilt$BGN_DATE<-strptime(stormfilt$BGN_DATE,format="%m/%d/%Y")
    stormfilt$BGN_DATE<-as.POSIXct(stormfilt$BGN_DATE)
    stormfilt$EVTYPE<-toupper(stormfilt$EVTYPE)
    stormfilt$Year<-format(stormfilt$BGN_DATE,format="%Y")
    stormfilt<-stormfilt[stormfilt$EVTYPE!="?",]
```

Now I am going to download the list of event types (EVTYPES) from the following
link. Using this list I will then match up the official event types with those
in the storm data set. These event types are listed on page 6 of the document
under "2.1.1 Storm Data Event Table". The locate areas function brings up a UI
showing the sixth page of the pdf which the user can use to select the area around
the table of text. The extract tables function will then use these dimensions in
the area argument to pull items from there. I have saved the dimensions of the table
in a list for the sake of reproducibility.
```{r Clean Event Types, cache=TRUE}
fileurl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
#x<-locate_areas(fileurl,pages=6)
tablelocation<-list(c(197.6,67.6,564.6,524))
stormeventtypes<-extract_tables(fileurl,output="data.frame",pages=6,
                                area=tablelocation)
stormET<-c(stormeventtypes[[1]][,1],stormeventtypes[[1]][,3])
#Remove entries that are not event types
stormET<-stormET[-c(1,2,27,28)]
stormET<-toupper(stormET)
#used jaro-winkler distance (jw) since there are many entries in which there are
#abbreviations and extra words unrelated to categories. This method worked better
#than the default, and resolved NA matches that occurred with the default
stormfilt$EVTYPECOND<-stormET[amatch(stormfilt$EVTYPE,stormET,method="jw",maxDist = 7)]
stormfilt$EVTYPECOND<-as.factor(stormfilt$EVTYPECOND)
```

The next step is to use the CROPDMGEXP and PROPDMGEXP to adjust the actual crop
and property damage totals. There are a few entries in these that only change the 
damage value by a trivial amount (hundreds of dollars) and these will be removed
from the data set, since they will not have a large impact on the analysis as 
damage is typically in the millions or billions. These entries also only account
for a couple hundred observations.
```{r Update damage costs}
stormfilt$PROPDMGEXP <- toupper(stormfilt$PROPDMGEXP) 
stormfilt$CROPDMGEXP <- toupper(stormfilt$CROPDMGEXP)
#Instances of *EXP I am looking to keep
modifier<-c("M","K","B","H","")
stormfilt<-filter(stormfilt,CROPDMGEXP %in% modifier & PROPDMGEXP %in% modifier)
stormfilt<-stormfilt %>%
                mutate(.,PROPDMGFULL= with(.,case_when(
                    (PROPDMGEXP=="H") ~ PROPDMG*100,
                    (PROPDMGEXP=="K") ~ PROPDMG*1000,
                    (PROPDMGEXP=="M") ~ PROPDMG*10^6,
                    (PROPDMGEXP=="B") ~ PROPDMG*10^9,
                    TRUE ~ PROPDMG
                )))
                          
stormfilt<-stormfilt %>%
                mutate(.,CROPDMGFULL= with(.,case_when(
                    (CROPDMGEXP=="H") ~ CROPDMG*100,
                    (CROPDMGEXP=="K") ~ CROPDMG*1000,
                    (CROPDMGEXP=="M") ~ CROPDMG*10^6,
                    (CROPDMGEXP=="B") ~ CROPDMG*10^9,
                    TRUE ~ CROPDMG
                )))
```

### Data Selection-Population Health

The first question to be addressed is which types of weather events are most 
harmful to population health.There are two variables in the data set, Fatalities 
and Injuries which will act as measures of population health. I will begin the 
analysis by examining the total fatalities and injuries caused by various weather 
event types. I am using total because this gives the best indication of the 
severity of the weather event. This will then help to filter out which event 
types are of the most interest.

I am also only going to examine data from 1996 and on, since this is when all 
categories of event types were collected. Otherwise the data would be extremely
skewed to reflect data on event types that have been collected since the 
beginning (e.g. Tornadoes)

```{r Population health analysis}
totalstats<-stormfilt %>%
    filter(Year>"1995" & (FATALITIES>0 | INJURIES>0 )) %>%
    group_by(EVTYPECOND) %>%
    summarise(Total_Fatalities=sum(FATALITIES),Total_Injuries=sum(INJURIES))
```


Now to examine the most signficant weather events, lets use event types that are
above the 75 percentile for total fatalities and injuries

```{r quantile data selection}
#Establishing quantle cutoff values
quant<-totalstats %>%
        summarise(quantFAT=quantile(Total_Fatalities,probs=.75),
              quantINJ=quantile(Total_Injuries,probs=.75))

Eventnames<-totalstats$EVTYPECOND[totalstats$Total_Fatalities>quant$quantFAT & 
               totalstats$Total_Injuries>quant$quantINJ]

healthdata<-totalstats%>%
    filter(EVTYPECOND %in% Eventnames)%>%
    arrange(desc(Total_Fatalities))
    
healthdata$EVTYPECOND<-factor(healthdata$EVTYPECOND,levels=as.character(healthdata$EVTYPECOND))

```

### Data Selection-Economic Consequences of Weather Events

Now I will do an analysis focusing on the economic impact of weather events. 
This will be very similar in scope to the previous analysis, focusing on events
after 1995 and including events that caused either crop damage, or property damage.
Then I will select events that caused the most property damage as selected by the 
75th percentile.
```{r Population economic analysis}
totalcost<-stormfilt %>%
    filter(Year>"1995" & (PROPDMGFULL>0 | CROPDMGFULL>0 )) %>%
    group_by(EVTYPECOND) %>%
    summarise(Total_Property_Damage=sum(PROPDMGFULL),
              Total_Crop_Damage=sum(CROPDMGFULL))

#Establishing quantle cutoff values
quantcost<-totalcost %>%
        summarise(quantPROP=quantile(Total_Property_Damage,probs=.75),
              quantCROP=quantile(Total_Crop_Damage,probs=.75))

Eventnames<-totalcost$EVTYPECOND[totalcost$Total_Property_Damage>quantcost$quantPROP & 
               totalcost$Total_Crop_Damage>quantcost$quantCROP]

costdata<-totalcost%>%
    filter(EVTYPECOND %in% Eventnames)%>%
    arrange(desc(Total_Property_Damage))
    
costdata$EVTYPECOND<-factor(costdata$EVTYPECOND,levels=as.character(costdata$EVTYPECOND))

```

# Results

```{r Tidytext function, echo=FALSE}
#### This function is part of the TIDYTEXT package authored by Julia Silge
#### and David Robinson. The package was not downloading this function
#### so I used the code they have documented in the function
#### https://rdrr.io/github/juliasilge/tidytext/src/R/reorder_within.R

reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  if (!is.list(within)) {
    within <- list(within)
  }

  new_x <- do.call(paste, c(list(x, sep = sep), within))
  stats::reorder(new_x, by, FUN = fun)
}

```

### Plots Demonstrating harmful effects of Severe Weather events on Population Health

```{r Health plots}
healthdata %>%
     gather(key,value,-EVTYPECOND) %>%
    ggplot(aes(x=reorder_within(EVTYPECOND,-value,key),y=value,
                           fill=EVTYPECOND))+
     geom_bar(stat="identity",position="dodge")+
    facet_wrap(~key,scales="free")+
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.x = element_blank(),
          plot.title = element_text(hjust=.5),
          legend.position = "bottom")+
    scale_fill_brewer(palette = "Set1")+
    labs(y="Total Number of Incidents",
         title = "Total Fatalities and Injuries due to Weather from 1996-2011")+
    guides(fill=guide_legend(title="Weather Event"))

```

#### Plot Summary

These plots show the total amount of fatalities and injuries as a result of 8 
weather events from 1996-2011. These weather events were chosen as they were in
the 75th percentile of fatalities and injuries. Notice the scale between 
fatalities and injuries differs by abouta factor of 10 (Many more injuries than fatalities). 
Examination of the Fatalities plot shows that Excessive Heat, and Tornado are the 
largest contributors to fatalities, resulting in more than one thousand fatalities. 
In the total injuries plot, tornado, flood, and excessive heat produce that largest number of injuries.Based on this, Tornadoes seem to be the most harmful to population health as they
result in large amounts of fatalities and massive amounts of injuries. Excessive
heat is the next largest threat to health causing a large amount of fatalities and
injuries. The hierarchy of weather events from here is less clear, but all the
other weather events result in hundreds of fatalities and thousands of injuries.

### Plots demonstrating harmful economic effects of Severe Weather events 

```{r Cost plots}
costdata %>%
     gather(key,value,-EVTYPECOND) %>%
    ggplot(aes(x=reorder_within(EVTYPECOND,-value,key),y=value,
                           fill=EVTYPECOND))+
     geom_bar(stat="identity",position="dodge")+
    facet_wrap(~key,scales="free")+
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.x = element_blank(),
          plot.title = element_text(hjust=.5),
          legend.position = "bottom")+
    scale_fill_brewer(palette = "Set2")+
    labs(y="Total Cost (in Dollars)",
         title = "Total Crop and Property Damage due to Weather from 1996-2011")+
    guides(fill=guide_legend(title="Weather Event"))

```

### Plot Summary
This figure contains two plots, one showing the total crop damage and the other total
property damage as a result of the most damaging weather events between 1996 and 2011.
Notice that the scales of total crop damage and total property damage are very different,
with total crop damage maxing out at slightly less than 6 billion dollars, whereas
total property damage maxes out at slightly less than 150 billion dollars. 

#### Total Crop Damage

In relation to total crop damage, hurricanes (typhoon) caused the largest amount of total
crop damage resulting in approximately 5.5 billion dollars worth of damage. Floods
were slightly less than this with about 5 billion dollars in damage. Hail was the
next most crop damaging weather event causing about 2.5 billion dollars worth of
damage. Flash floods and high winds were the next most damaging events causing a 
little over 1 billion dollars in crop damage. Finally tropical storms and wildfires
were next resulting in hundreds of millions dollars worth of crop damage.

#### Total Property Damage

In relation to total property damage, Floods resulted in slightly less than 150
billion dollars worth of damage. The next most damaging weather event was 
hurricane (typhoon) coming in at around 80 billion dollars worth of property damage.
Flash floods and hail were the next most damaging weather events resulting in 
approximately 15 billion dollars worth of damage. High winds caused almost 10 billion
dollars worth of damage, while tropical storms and wildfires resulted in 7.5 
billion dollars worth of damage.

### Table of Economic and Population Health Consequences of Weather Events
For the interested reader who wants exact values for the variables of interest,
I have also included this table which shows values for all weather events.
``` {r health and economic cost table}
alltotal<-merge(totalstats,totalcost,by="EVTYPECOND",all=TRUE)
    datatable(alltotal,class="cell-border stripe")

```


This table shows the weather events that result in damage to population 
health or the economy between 1996 and 2011.
